---
title: "readstata13: Read and write the 'Stata' file format with R"
output:
  github_document:
    fig_width: 4
    fig_height: 3.5
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-")

options(knitr.kable.NA = '')
```

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/readstata13)](https://cran.r-project.org/package=readstata13)
[![Build status](https://github.com/sjewo/readstata13/workflows/R-CMD-check/badge.svg)](https://github.com/sjewo/readstata13/actions?workflow=R-CMD-check)
[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/readstata13)](https://cran.r-project.org/package=readstata13)
<!-- badges: end -->


Package to read and write all Stata file formats (version 17 and older) into a
R data.frame. The dta file format versions 102 to 121 are supported (including dtas files).

The function ```read.dta``` from the foreign package imports only dta files from
Stata versions <= 12. Due to the different structure and features of dta 117
files, we wrote a new file reader in Rcpp.

Additionally the package supports many features of the Stata dta format like
label sets in different languages (`?set.lang`) or business calendars
(`?as.caldays`).


## Installation

The package is hosted on CRAN.
```{r, eval = FALSE}
install.packages("readstata13")
```

## Usage
```{r, eval = FALSE}
library(readstata13)
dat <- read.dta13("path to file.dta")
save.dta13(dat, file="newfile.dta")
```

## Development Version
To install the current release from github you need the platform specific build
tools. On Windows a current installation of
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) is necessary, while OS X
users need to install
[Xcode](https://apps.apple.com/us/app/xcode/id497799835).

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("sjewo/readstata13", ref="0.11")
```

To install the current development version from github:

```{r, eval = FALSE}
remotes::install_github("sjewo/readstata13", ref="testing")
```


## Changelog and Features

```{r, echo = FALSE}
news <- readLines("NEWS")
news_stop <- which(news == news[grepl("^\\[", news)][2]) - 2
news <- news[1:news_stop]

news_tab <- data.frame(Version = gsub("\\[|\\]", "", news[1]), 
                       Changes = gsub("^- ", "", news[-1]))
news_tab[2:nrow(news_tab), "Version"] <- NA
knitr::kable(news_tab)
```

See [News](NEWS) for the full changelog.

```{r, echo = FALSE}
# create NEWS.MD for pagedown

news <- readLines("NEWS")
newsmd <- gsub("^\\[(.*)\\]", "## readstata13 \\1 \n", news)
cat(newsmd, file ="NEWS.md", sep = "\n")
```


## readstata13 and foreign

Most attributes of the resulting data.frame are largely similar to the data.frames produced by `foreign`. 
Since newer Stata files require some additional attributes, the results of `all.equal()` and `identical()` will be `FALSE` for data.frames read by `foreign::read.dta` and `read.dta13()`.
Otherwise, the data.frames produced by both functions are identical.

```{r, eval = FALSE}
library(foreign)
library(readstata13)

# with factors
r12 <- read.dta("http://www.stata-press.com/data/r12/auto.dta")
r13 <- read.dta13("http://www.stata-press.com/data/r13/auto.dta")

all.equal(r12, r13, check.attributes = FALSE)

# without factors
r12 <- read.dta("http://www.stata-press.com/data/r12/auto.dta", 
                convert.factors = FALSE)
r13 <- read.dta13("http://www.stata-press.com/data/r13/auto.dta", 
                  convert.factors = FALSE)

all.equal(r12, r13, check.attributes = FALSE)
```

## Authors

[Marvin Garbuszus](mailto:jan.garbuszus@ruhr-uni-bochum.de) ([JanMarvin](https://github.com/JanMarvin)) and [Sebastian Jeworutzki](mailto:Sebastian.Jeworutzki@ruhr-uni-bochum.de) ([sjewo](https://github.com/sjewo)) 

## Licence

GPL2
